---
title: "Introduction to ScIGMA"
author: 
  - name: "Geoffrey Pawlak"
    affiliation: "Systems Oncology & Bioinformatics"
date: "`r Sys.Date()`"
output: BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{Introduction to ScIGMA}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

**ScIGMA** (Single-cell Integrated Genomic & Multi-omics Analyzer) is a comprehensive computational framework for exploring single-cell proteogenomic architecture. This vignette demonstrates the basic usage of the ScIGMA interactive application.

## Launching the Application

ScIGMA is primarily designed to be run as an interactive Shiny application. Once the package is installed, you can launch the interface using the `run_app()` function.

```{r setup, eval=FALSE}
library(ScIGMA)

# Launch the interactive interface
ScIGMA::run_app()
```

## Data Input Requirements

The application expects high-dimensional single-cell data, primarily targeting `.h5` objects generated from systems like Mission Bio Tapestri. 
The expected matrices include:
- Single Nucleotide Variants (SNV) matrices (Genotype, Depth, Allele Frequency)
- Copy Number Variations (CNV) matrices
- Protein Expression matrices

## Downstream Analysis

Once data is loaded into the interactive interface, ScIGMA allows for:
1. **Filtering**: Removal of low-quality cells and non-informative variants.
2. **Inference**: Imputation of missing genotypes and inference of clonal architecture using the underlying COMPASS C++ engine via MCMC.
3. **Annotation**: Direct mapping of clinical significance and variant consequences via Ensembl VEP API.
4. **Integration**: UMAP dimensionality reduction of the protein space and manual/automated gating of immunophenotypes.

# Session Information

```{r sessionInfo}
sessionInfo()
```
